Efficient analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2018-02-02 cabios-template2/bigsnpr-review.tex
replace suppmat refs

\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{../figures/}}
\usepackage{xr}
\externaldocument{bigsnpr-review-supp}

@@

\title[R packages for analyzing genome-wide data]{Efficient analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author[Sample \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$, Andrey Ziyatdinov\,$^{\text{\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Univ.\ Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France, \\
$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}

@@

\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. {\color{red} This increase in data size is a major challenge severely slowing down genomic analyses, leading to some software becoming obsolete and researchers having limited access to diverse analysis tools.} \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for the analysis of large scale genomic data to be performed within R. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement fast and accurate computations of principal component analysis and association studies, functions to remove SNPs in linkage disequilibrium and algorithms to learn polygenic risk scores on millions of SNPs. We illustrate applications of the two R packages by analyzing a case-control genomic dataset for celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} \& \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary materials are available at \textit{Bioinformatics}
online.}

@@

{\color{red}
Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
As a consequence, most existing software and algorithms have to be continuously optimized in order to avoid obsolescence. For computing Principal Component Analysis (PCA), commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to the software EIGENSOFT, and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}. 
PLINK 1.07, which has been a central tool in the analysis of genotype data, has been replaced by PLINK 1.9 to speed-up computations, and there is also an alpha version of PLINK 2.0 that will handle more data types \cite[]{chang2015second,Purcell2007}.

Increasing size of genetic datasets is a source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers.
This is particularly a burden for commonly used analysis languages such as R.
For analyzing genotype datasets in R, a range of software are available, including for example the popular R packages GenABEL, SNPRelate and GWASTools \cite[]{aulchenko2007genabel,Gogarten2012,zheng2012high}. 
Solving memory issues for languages such as R would give access to a broad range of already implemented tools for data analysis. Fortunately, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
}

@@

{\color{red}
In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
R is a programming language that makes it easy to tie together existing or new functions to be used as part of large, interactive and reproducible analyses \cite[]{R2017}.
We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data. This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.
}
Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double). This includes implementation of multivariate sparse linear models, principal component analysis, association tests, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting). 
{\color{red}
We use both a real case-control genomic dataset for celiac disease and large-scale simulated data to illustrate application of the two R packages, including two association studies and the computation of Polygenic Risk Scores (PRS). We compare results from bigstatsr and bigsnpr with those obtained by using command-line sofwtare PLINK, EIGENSOFT and PRSice, and R packages SNPRelate and GWASTools. We report execution times along with the code to perform major computational tasks. 
For a comprehensive comparison between R packages bigstatsr and bigmemory, see supplementary notebook "bigstatsr-and-bigmemory".
}

@@

{\color{red}
We developed a special FBM object, called ``FBM.code256'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage. Indeed, each element is stored in one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed'' which can store only genotype calls. With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values). So, we use a single data format that can store both genotype calls and dosages.

For preprocessing steps, PLINK is a widely-used software. For the sake of reproducibility, one could use PLINK directly from R via systems calls. We therefore provide wrappers as R functions that use system calls to PLINK for conversion and quality control and a variety of formats can be used as input (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
Package bigsnpr provides fast conversions between bed/bim/fam PLINK files and the ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM.code256), a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''. Finally, we provide two methods for converting dosage data to the format ``bigSNP'' (Supplementary notebook ``dosage'').

Most modern SNP chips provide genotype data with large call-rates. For example, the Celiac data we use in this paper presents only 0.04\% of missing values after quality control. Yet, most of the functions in bigstatsr and bigsnpr do not handle missing values.
So, we provide two functions for imputing missing values of genotyped SNPs. Note that we do not impute completely missing SNPs which would require the use of reference panels and could be performed via e.g.\ imputation servers for human data \cite[]{mccarthy2016reference}. 
The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R (Figure~\ref{fig:impute}). 
The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. This function also provides an estimator of the imputation error rate by SNP for post quality control.
This algorithm is based on Machine Learning approaches for genetic imputation \cite[]{Wang2012} and does not use phasing, thus allowing for a dramatic decrease in computation time. It only relies on some local XGBoost models \cite[]{Chen2016}. 
XGBoost, which is available in R, builds decision trees that can detect nonlinear interactions, partially reconstructing phase, making it well suited for imputing genotype matrices. 
Our algorithm is the following: for each SNP, we divide the individuals in the ones which have a missing genotype (test set) and the ones which have a non-missing genotype for this particular SNP. Those latter individuals are further separated in a training set and a validation set (e.g.\ 80\% training and 20\% validation). The training set is used to build the XGBoost model for predicting missing data. The prediction model is then evaluated on the validation set for which we know the true genotype values, providing an estimator of the number of genotypes that have been wrongly imputed for that particular SNP. The prediction model is also projected on the test set (missing values) in order to impute them.
}

@@

SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}. There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium. Two of them, named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}. 
First, pruning is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
{\color{red}
Clumping is useful if a statistic is available for sorting the SNPs by importance. Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome. 
For PCA, the thinning procedure should remain unsupervised (no phenotype must be used) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance. 
This choice is consistent with the pruning algorithm of PLINK; when two nearby SNPs are correlated, PLINK keeps only the one with the highest MAF.
Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all (Supplementary notebook ``pruning-vs-clumping''). 
So, we suggest to use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr. In practice, for the three datasets we considered, the clumping algorithm with the MAF provides similar sets of SNPs as when using the pruning algorithm (results not shown).
}

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcomes in bigsnpr \cite[]{Zheng2012}. 
Package bigstatsr implements statistical tests based on linear and logistic regressions. For linear regression, a t-test is performed for each SNP $j$ on $\beta^{(j)}$ where
\begin{multline}
  \hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $K$ is the number of principal components and $L$ is the number of other covariates (such as age and gender). Similarly, for logistic regression, a Z-test is performed for each SNP $j$ on $\beta^{(j)}$ where
\begin{multline}
  \log{\left(\frac{\hat{p}}{1-\hat{p}}\right)} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
{\color{red}
These tests can be used to perform genome-wide association studies (GWAS) and are very fast due to the use of optimized implementations, partly based on previous work by \cite{sikorska2013gwas}. 
}
%For a continuous phenotype $y$ and a SNP $s$ and some covariables $X$ (including a column of 1s for the intercept), we can compute $s^* = s - X(X^TX)^{-1} X^T s$ and $y^* = s - X(X^TX)^{-1} X^T y$ (orthogonal projections relative to the space of covariables).

The R packages also implement functions for computing Polygenic Risk Scores using two methods. 
{\color{red} The first method is the widely-used ``Clumping + Thresholding'' (C+T, also called ``Pruning + Thresholding'' in the literature) model based on univariate GWAS summary statistics as described in previous equations.}
Under the C+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). The SNPs are first clumped (C) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). A polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Chatterjee2013,Dudbridge2013,Euesden2015}. 
On the contrary, the second approach does not use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables \textit{at once}, optimally accounting for correlation between predictors \cite[]{Abraham2012}. The currently available models are very fast sparse linear and logistic regressions. These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Friedman2010,Tibshirani1996,Zou2005}. Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}. The implementation of these algorithms is based on modified versions of functions available in the R package biglasso \cite[]{Zeng2017}. These modifications allow to include covariates in the models, to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr and to remove the need of choosing the regularization parameter.

@@

In this paper, two datasets are analyzed: the celiac disease cohort and POPRES \cite[]{Dubois2010,Nelson2008}. The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
For computation time comparisons, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same {\color{red} eigen decomposition (up to a constant) and pairwise SNP correlations} as the original dataset. To assess scalibility of the packages for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset. \label{sec:rep}

@@

All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the supplementary materials. 
{\color{red}
In supplementary notebook ``public-data'', we provide some open-access data of domestic dogs so that users can test our code and functions on a moderate size dataset with 4342 samples and 145,596 SNPs \cite[]{hayward2016complex}.
}

@@

We present the results of four different analyses. 
First, we illustrate the application of R packages bigstatsr and bigsnpr. 
Secondly, by performing two GWAS, we compare the performance of bigstatsr and bigsnpr to the performance obtained with FastPCA (EIGENSOFT 6.1.4) and PLINK 1.9, and also two R packages SNPRelate and GWASTools \cite[]{chang2015second,Galinsky2016,Gogarten2012,zheng2012high}. PCA is a computationally intensive step of the GWAS, so that we further compare PCA methods on larger datasets.
Thirdly, by performing a PRS analysis with summary statistics, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (i.e.\ for genotyped SNPs only). 
We compare performance on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores). For the functions that enable parallelism, we use half of the cores available on the corresponding computer. We present a table summarizing the features of different software in table \ref{tab:andrey}.

@@

The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs with more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency (MAF) lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}. 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\%. The 0.04\% missing genotype values were imputed with the XGBoost method. If we would have used a standard R matrix to store the genotypes, this data would have required 32GB of memory. On the disk, the ``.bed'' file requires 1GB and the ``.bk'' file (storing the FBM) requires 4GB.

@@

For detecting long-range LD regions during the computation of PCA, we tested the function snp\_autoSVD on both the Celiac and POPRES datasets. For the POPRES dataset, the algorithm converged in two iterations. The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table~\ref{tab:lrldr-popres}). 
We compared the PCs of genotypes obtained after applying snp\_autoSVD with the PCs obtained after removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}). 
For the Celiac dataset, we found 5 long-range LD regions (Table~\ref{tab:lrldr-celiac}) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping and removing predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).

>>>>

\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
%\graphicspath{{../figures/}}
%\usepackage{xr}
%\externaldocument{bigsnpr-review-supp}

@@

\title[R packages for analyzing genome-wide data]{Efficient analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author[Priv\'e \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$, Andrey Ziyatdinov\,$^{\text{\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Univ.\ Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France, \\
$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}

@@

\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge severely slowing down genomic analyses, leading to some software becoming obsolete and researchers having limited access to diverse analysis tools. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for the analysis of large scale genomic data to be performed within R. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement fast and accurate computations of principal component analysis and association studies, functions to remove SNPs in linkage disequilibrium and algorithms to learn polygenic risk scores on millions of SNPs. We illustrate applications of the two R packages by analyzing a case-control genomic dataset for celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} \& \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary materials are available at \textit{Bioinformatics}
online.}

@@

Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
As a consequence, most existing software and algorithms have to be continuously optimized in order to avoid obsolescence. For computing Principal Component Analysis (PCA), commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to the software EIGENSOFT, and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}. 
PLINK 1.07, which has been a central tool in the analysis of genotype data, has been replaced by PLINK 1.9 to speed-up computations, and there is also an alpha version of PLINK 2.0 that will handle more data types \cite[]{chang2015second,Purcell2007}.

Increasing size of genetic datasets is a source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers.
This is particularly a burden for commonly used analysis languages such as R.
For analyzing genotype datasets in R, a range of software are available, including for example the popular R packages GenABEL, SNPRelate and GWASTools \cite[]{aulchenko2007genabel,Gogarten2012,zheng2012high}. 
Solving memory issues for languages such as R would give access to a broad range of already implemented tools for data analysis. Fortunately, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.

@@

In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
R is a programming language that makes it easy to tie together existing or new functions to be used as part of large, interactive and reproducible analyses \cite[]{R2017}.
We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data. This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.
Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double). This includes implementation of multivariate sparse linear models, principal component analysis, association tests, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting). 
We use both a real case-control genomic dataset for celiac disease and large-scale simulated data to illustrate application of the two R packages, including two association studies and the computation of Polygenic Risk Scores (PRS). We compare results from bigstatsr and bigsnpr with those obtained by using command-line sofwtare PLINK, EIGENSOFT and PRSice, and R packages SNPRelate and GWASTools. We report execution times along with the code to perform major computational tasks. 
For a comprehensive comparison between R packages bigstatsr and bigmemory, see supplementary notebook "bigstatsr-and-bigmemory".

@@

We developed a special FBM object, called ``FBM.code256'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage. Indeed, each element is stored in one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed'' which can store only genotype calls. With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values). So, we use a single data format that can store both genotype calls and dosages.

For preprocessing steps, PLINK is a widely-used software. For the sake of reproducibility, one could use PLINK directly from R via systems calls. We therefore provide wrappers as R functions that use system calls to PLINK for conversion and quality control and a variety of formats can be used as input (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~S1).
Package bigsnpr provides fast conversions between bed/bim/fam PLINK files and the ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM.code256), a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''. Finally, we provide two methods for converting dosage data to the format ``bigSNP'' (Supplementary notebook ``dosage'').

Most modern SNP chips provide genotype data with large call-rates. For example, the Celiac data we use in this paper presents only 0.04\% of missing values after quality control. Yet, most of the functions in bigstatsr and bigsnpr do not handle missing values.
So, we provide two functions for imputing missing values of genotyped SNPs. Note that we do not impute completely missing SNPs which would require the use of reference panels and could be performed via e.g.\ imputation servers for human data \cite[]{mccarthy2016reference}. 
The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R (Figure~S2). 
The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. This function also provides an estimator of the imputation error rate by SNP for post quality control.
This algorithm is based on Machine Learning approaches for genetic imputation \cite[]{Wang2012} and does not use phasing, thus allowing for a dramatic decrease in computation time. It only relies on some local XGBoost models \cite[]{Chen2016}. 
XGBoost, which is available in R, builds decision trees that can detect nonlinear interactions, partially reconstructing phase, making it well suited for imputing genotype matrices. 
Our algorithm is the following: for each SNP, we divide the individuals in the ones which have a missing genotype (test set) and the ones which have a non-missing genotype for this particular SNP. Those latter individuals are further separated in a training set and a validation set (e.g.\ 80\% training and 20\% validation). The training set is used to build the XGBoost model for predicting missing data. The prediction model is then evaluated on the validation set for which we know the true genotype values, providing an estimator of the number of genotypes that have been wrongly imputed for that particular SNP. The prediction model is also projected on the test set (missing values) in order to impute them.

@@

SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}. There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium. Two of them, named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}. 
First, pruning is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
Clumping is useful if a statistic is available for sorting the SNPs by importance. Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome. 
For PCA, the thinning procedure should remain unsupervised (no phenotype must be used) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance. 
This choice is consistent with the pruning algorithm of PLINK; when two nearby SNPs are correlated, PLINK keeps only the one with the highest MAF.
Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all (Supplementary notebook ``pruning-vs-clumping''). 
So, we suggest to use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr. In practice, for the three datasets we considered, the clumping algorithm with the MAF provides similar sets of SNPs as when using the pruning algorithm (results not shown).

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcomes in bigsnpr \cite[]{Zheng2012}. 
Package bigstatsr implements statistical tests based on linear and logistic regressions. For linear regression, a t-test is performed for each SNP $j$ on $\beta^{(j)}$ where
\begin{multline}
  \hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $K$ is the number of principal components and $L$ is the number of other covariates (such as age and gender). Similarly, for logistic regression, a Z-test is performed for each SNP $j$ on $\beta^{(j)}$ where
\begin{multline}
  \log{\left(\frac{\hat{p}}{1-\hat{p}}\right)} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
These tests can be used to perform genome-wide association studies (GWAS) and are very fast due to the use of optimized implementations, partly based on previous work by \cite{sikorska2013gwas}.

The R packages also implement functions for computing Polygenic Risk Scores using two methods. 
The first method is the widely-used ``Clumping + Thresholding'' (C+T, also called ``Pruning + Thresholding'' in the literature) model based on univariate GWAS summary statistics as described in previous equations.
Under the C+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). The SNPs are first clumped (C) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). A polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Chatterjee2013,Dudbridge2013,Euesden2015}. 
On the contrary, the second approach does not use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables \textit{at once}, optimally accounting for correlation between predictors \cite[]{Abraham2012}. The currently available models are very fast sparse linear and logistic regressions. These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Friedman2010,Tibshirani1996,Zou2005}. Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}. The implementation of these algorithms is based on modified versions of functions available in the R package biglasso \cite[]{Zeng2017}. These modifications allow to include covariates in the models, to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr and to remove the need of choosing the regularization parameter.

@@

In this paper, two datasets are analyzed: the celiac disease cohort and POPRES \cite[]{Dubois2010,Nelson2008}. The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
For computation time comparisons, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same eigen decomposition (up to a constant) and pairwise SNP correlations as the original dataset. To assess scalibility of the packages for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset. \label{sec:rep}

@@

All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the supplementary materials. 
In supplementary notebook ``public-data'', we provide some open-access data of domestic dogs so that users can test our code and functions on a moderate size dataset with 4342 samples and 145,596 SNPs \cite[]{hayward2016complex}.

@@

We present the results of four different analyses. 
First, we illustrate the application of R packages bigstatsr and bigsnpr. 
Secondly, by performing two GWAS, we compare the performance of bigstatsr and bigsnpr to the performance obtained with FastPCA (EIGENSOFT 6.1.4) and PLINK 1.9, and also two R packages SNPRelate and GWASTools \cite[]{chang2015second,Galinsky2016,Gogarten2012,zheng2012high}. PCA is a computationally intensive step of the GWAS, so that we further compare PCA methods on larger datasets.
Thirdly, by performing a PRS analysis with summary statistics, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (i.e.\ for genotyped SNPs only). 
We compare performance on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores). For the functions that enable parallelism, we use half of the cores available on the corresponding computer. We present a table summarizing the features of different software in table S9.

@@

The data was preprocessed following steps from figure~S1, removing individuals and SNPs with more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency (MAF) lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}. 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\%. The 0.04\% missing genotype values were imputed with the XGBoost method. If we would have used a standard R matrix to store the genotypes, this data would have required 32GB of memory. On the disk, the ``.bed'' file requires 1GB and the ``.bk'' file (storing the FBM) requires 4GB.

@@

For detecting long-range LD regions during the computation of PCA, we tested the function snp\_autoSVD on both the Celiac and POPRES datasets. For the POPRES dataset, the algorithm converged in two iterations. The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table~S1). 
We compared the PCs of genotypes obtained after applying snp\_autoSVD with the PCs obtained after removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~S2). 
For the Celiac dataset, we found 5 long-range LD regions (Table~S3) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping and removing predetermined long-range LD regions (Table~S4).

For the Celiac dataset, we further compared results of PCA obtained when using snp\_autoSVD and when computing PCA without removing any long range LD region (only clumping at $R^2 > 0.2$). 
When not removing any long range LD region, we show that PC4 and PC5 do not capture population structure and correspond to a long-range LD region in chromosome 8 (Figures~S3 and~S4). 
When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 reflect population structure (Figure~S3). Moreover, loadings are more equally distributed among SNPs after removal of long-range LD regions (Figure~S4). 
This is confirmed by Gini coefficients (measure of dispersion) of each squared loadings that are significantly smaller when computing PCA with snp\_autoSVD than when no long-range LD region is removed (Figure~S5).
